---
title: "S01_analysis"
author: "Cal Faubel & Kyle Hilliam"
output: html_document
date: "2026-03-13"
editor_options:
  chunk_output_type: console
---

Code developed using R version 4.3 
Remote code repository location: https://github.com/CalFaubel/Caulerpa_analysis 

Project title: 
Network analysis of vessel movements facilitates rapid response to invasive marine species incursion

Purpose: 
This code outlines the analysis as part of the publication listed above. 
Here we show the process of creating networks from a node and edge list to export as shapefile for visualistaion in GIS software along with figure creation.  
Please contact Cal Faubel or Kyle Hilliam for querries related to this code.
- c.faubel@aims.gov.au or calfaubel33@gmail.com
- kyle.hilliam@outlook.com

Input data description: 
Data imported into this analysis can be found here: https://github.com/CalFaubel/Caulerpa_analysis/data/


# Aim: Undertake Network Analysis 

# Labelling System
- .df, .list, .dist etc at the end of an object signify what class it is
- df with spatial information and sf or sp objects will contain the coordinate system in their name, e.g., buffer_WGS1984.sf once transformed. But will not in first instance (buffer.df)

  
# 1. Load packages 
```{r}
library(sf)
library(tidyverse)
library(dplyr)
library(igraph)
library(readr)
library(ggtext)
```

```{r Identify Machine}
if (Sys.info()[["nodename"]] == "station") {
  setwd("/home/khill/GitHub/Caulerpa_analysis/")
}
```

# 2. Load data 
Two vessel datasets 
- Automated Identification System (AIS)
- Survey data set 

Both split into 
1. Downstream: vessel movements 10 days after visiting Cauelrpa sites visited
2. Reachability: vessel movements after visiting Cauelrpa sites, no constraints.
```{r}
Downstream_AIS_edge.df       <- read_csv("data/Downstream_AIS_edge.csv")

Downstream_AIS_nodes.df      <- read_csv("data/Downstream_AIS_nodes.csv")

Downstream_survey_edge.df    <- read_csv("data/Downstream_survey_edge.csv")

Downstream_survey_nodes.df   <- read_csv("data/Downstream_survey_nodes.csv")

Reachability_AIS_edge.df     <- read_csv("data/Reachability_AIS_edge.csv")

Reachability_AIS_nodes.df    <- read_csv("data/Reachability_AIS_nodes.csv")

Reachability_survey_edge.df  <- read_csv("data/Reachability_survey_edge.csv")

Reachability_survey_nodes.df <- read_csv("data/Reachability_survey_nodes.csv")
```


# 3. Create igraph object and calcualte metrics 

## 3.1 Create Igraph object 

### 3.1.1 Downstream_AIS
create I graph objects. 
- Important to set your weight attribute 
```{r}
head(Downstream_AIS_edge.df)

AIS_down.g <- graph_from_data_frame(Downstream_AIS_edge.df, Downstream_AIS_nodes.df , directed = TRUE)

E(AIS_down.g)$weight    <- E(AIS_down.g)$vessels

```

#### Network Metrics 
Distance metrics:
Whilst not used in this analysis it is important to set your distance attribute that is used for metrics such as closeness centrality or Between centrality. this is set as the inverse of weight as larger weights should equate to shorter theoretical distances. 

Weighted degree measure:
see Opsahl et al., 2010 for how to set weight https://doi.org/10.1016/j.socnet.2010.03.006
```{r}
E(AIS_down.g)$strength <- E(AIS_down.g)$weight
E(AIS_down.g)$distance <- mean(E(AIS_down.g)$weight) / E(AIS_down.g)$weight  # define your distance as in your function. 

# --- Degree centrality ---
V(AIS_down.g)$degree    <- degree(AIS_down.g, v = V(AIS_down.g), mode = "total") 
V(AIS_down.g)$indegree  <- degree(AIS_down.g, v = V(AIS_down.g), mode = "in")
V(AIS_down.g)$outdegree <- degree(AIS_down.g, v = V(AIS_down.g), mode = "out")

# --- Strength
V(AIS_down.g)$strength     <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "all", loops = TRUE)
V(AIS_down.g)$in_strength  <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "in",  loops = TRUE)
V(AIS_down.g)$out_strength <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(AIS_down.g)$wei_degree     <- V(AIS_down.g)$degree * ((V(AIS_down.g)$strength / V(AIS_down.g)$degree)^0.5)
V(AIS_down.g)$wei_indegree   <- V(AIS_down.g)$indegree * ((V(AIS_down.g)$in_strength / V(AIS_down.g)$indegree)^0.5)
V(AIS_down.g)$wei_outdegree  <- V(AIS_down.g)$outdegree * ((V(AIS_down.g)$out_strength / V(AIS_down.g)$outdegree)^0.5)

```

### 3.1.2 Downstream_Survey
```{r}
head(Downstream_survey_edge.df)
head(Downstream_survey_nodes.df)

Survey_down.g <- graph_from_data_frame(Downstream_survey_edge.df, Downstream_survey_nodes.df , directed = TRUE)

E(Survey_down.g)$weight    <- E(Survey_down.g)$vessels

summary(Survey_down.g)
```

#### network metrics 
```{r}
E(Survey_down.g)$strength <- E(Survey_down.g)$weight
E(Survey_down.g)$distance <- mean(E(Survey_down.g)$weight) / E(Survey_down.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(Survey_down.g)$degree    <- degree(Survey_down.g, v = V(Survey_down.g), mode = "total")
V(Survey_down.g)$indegree  <- degree(Survey_down.g, v = V(Survey_down.g), mode = "in")
V(Survey_down.g)$outdegree <- degree(Survey_down.g, v = V(Survey_down.g), mode = "out")

# --- Strength 
V(Survey_down.g)$strength     <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "all", loops = TRUE)
V(Survey_down.g)$in_strength  <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "in",  loops = TRUE)
V(Survey_down.g)$out_strength <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(Survey_down.g)$wei_degree     <- V(Survey_down.g)$degree * ((V(Survey_down.g)$strength / V(Survey_down.g)$degree)^0.5) 
V(Survey_down.g)$wei_indegree   <- V(Survey_down.g)$indegree * ((V(Survey_down.g)$in_strength / V(Survey_down.g)$indegree)^0.5)
V(Survey_down.g)$wei_outdegree  <- V(Survey_down.g)$outdegree * ((V(Survey_down.g)$out_strength / V(Survey_down.g)$outdegree)^0.5)
```

### 3.1.3 Reachability_AIS
```{r}
head(Reachability_AIS_edge.df)

AIS_reach.g <- graph_from_data_frame(Reachability_AIS_edge.df, Reachability_AIS_nodes.df , directed = TRUE)

E(AIS_reach.g)$weight  <- E(AIS_reach.g)$vessels

summary(AIS_reach.g)
```

#### network metrics 
```{r}
E(AIS_reach.g)$strength <- E(AIS_reach.g)$weight
E(AIS_reach.g)$distance <- mean(E(AIS_reach.g)$weight) / E(AIS_reach.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(AIS_reach.g)$degree    <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "total")
V(AIS_reach.g)$indegree  <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "in")
V(AIS_reach.g)$outdegree <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "out")

# --- Strength 
V(AIS_reach.g)$strength     <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "all", loops = TRUE)
V(AIS_reach.g)$in_strength  <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "in",  loops = TRUE)
V(AIS_reach.g)$out_strength <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(AIS_reach.g)$wei_degree     <- V(AIS_reach.g)$degree * ((V(AIS_reach.g)$strength / V(AIS_reach.g)$degree)^0.5)
V(AIS_reach.g)$wei_indegree   <- V(AIS_reach.g)$indegree * ((V(AIS_reach.g)$in_strength / V(AIS_reach.g)$indegree)^0.5)
V(AIS_reach.g)$wei_outdegree  <- V(AIS_reach.g)$outdegree * ((V(AIS_reach.g)$out_strength / V(AIS_reach.g)$outdegree)^0.5)

```

### 3.1.4 Reachability_Survey
```{r}
head(Reachability_survey_edge.df)

Survey_reach.g <- graph_from_data_frame(Reachability_survey_edge.df, Reachability_survey_nodes.df , directed = TRUE)

E(Survey_reach.g)$weight  <- E(Survey_reach.g)$vessels

summary(Survey_reach.g)
```

#### network metrics 
```{r}
E(Survey_reach.g)$strength <- E(Survey_reach.g)$weight
E(Survey_reach.g)$distance <- mean(E(Survey_reach.g)$weight) / E(Survey_reach.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(Survey_reach.g)$degree    <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "total")
V(Survey_reach.g)$indegree  <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "in")
V(Survey_reach.g)$outdegree <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "out")

# --- Strength 
V(Survey_reach.g)$strength     <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "all", loops = TRUE)
V(Survey_reach.g)$in_strength  <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "in",  loops = TRUE)
V(Survey_reach.g)$out_strength <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(Survey_reach.g)$wei_degree     <- V(Survey_reach.g)$degree * ((V(Survey_reach.g)$strength / V(Survey_reach.g)$degree)^0.5)
V(Survey_reach.g)$wei_indegree   <- V(Survey_reach.g)$indegree * ((V(Survey_reach.g)$in_strength / V(Survey_reach.g)$indegree)^0.5)
V(Survey_reach.g)$wei_outdegree  <- V(Survey_reach.g)$outdegree * ((V(Survey_reach.g)$out_strength / V(Survey_reach.g)$outdegree)^0.5)

```

# 4 Exporting to shp 
use the igraph to shp function to export igraph object to shp file to be used for visualization in Arcpro 

## 4.1 Down_AIS 
```{r}
igraph2shp(AIS_down.g,shpOut = file.path(Network_Layer_Dir, 
                                            "AIS_down_MZTM"),
           CRSVal = "EPSG:2193")
```


## 4.2 Down Survey 
```{r}
igraph2shp(Survey_down.g,shpOut = file.path(Network_Layer_Dir, 
                                            "Survey_down_MZTM"),
           CRSVal = "EPSG:2193")
```

## 4.3 Reach AIS 
```{r}
igraph2shp(AIS_reach.g,shpOut = file.path(Network_Layer_Dir, 
                                            "AIS_reach_MZTM"),
           CRSVal = "EPSG:2193")
```

## 4.4 Reach Survey 
```{r}
igraph2shp(Survey_reach.g,shpOut = file.path(Network_Layer_Dir, 
                                            "Survey_reach_MZTM"),
           CRSVal = "EPSG:2193")
```


# 5. Priorty sites bar plot 

## 5.1 load in priorty sites csv 
```{r}
northland_priority_list.df <- read_csv("data/Northland_dodged_barplot_data.csv")
```

## 5.2 plot dodged bar plot 
```{r Dodged Bar Chart}
# Create empty row for nodes only in one set of data
single_data <- northland_priority_list.df %>%
  filter(n == 1) %>%
                mutate(per_inc = 0,
                       data = ifelse(data == "Survey", "AIS", ifelse(data == "AIS", "Survey", NA)))

# Bind together priority list with empty rows
northland_priority_list.df2 <- rbind(northland_priority_list.df, single_data) %>%
                                  arrange(nID, data) %>%
                                  group_by(nID) %>%
                                  mutate(n = n(),
                                         total_perc = sum(per_inc, na.rm = TRUE)) %>%
                                  ungroup()

# ggplot2 barchat with dodged bars for each node
dodged.barplot <- northland_priority_list.df2 %>%
                          ggplot(aes(x = reorder(Common_name_label, total_perc), y = per_inc, fill = data)) +
                          geom_bar(stat = "identity", position = position_dodge(width = 1)) +
                          labs(y = "Relative Risk (%)", x = "Locations") +
                          scale_fill_manual(name = "Node Data", values = c("#238b45", "#66c2a5")) +
                          scale_y_continuous(limits = c(0, 35), expand = c(0, 0)) +
                          theme_classic() +
                          theme(axis.text.y = element_markdown(size = 12, angle = 0, hjust = 1, margin = margin(t = 5, b = 5)), # Use element_markdown for bold formatting
                                axis.text.x = element_text(angle = 0),
                                axis.title = element_text(size = 14),
                                axis.text = element_text(size = 14),
                                legend.title = element_text(size = 14),
                                legend.text = element_text(size = 14)) +
                          coord_flip() 
                           

dodged.barplot

```


## 5.3 statistics

### 5.3.1 Correlation betwen incoming vessels 
Calculating the correation between the two datasets fro incoming vessels accross the 
downstream networks
```{r}
head(northland_priority_list.df)

# Pivot the data so that AIS and Survey per_inc values are side-by-side for each Common_name
correlation_df <- northland_priority_list.df %>%
  dplyr::select(Common_name, data, per_inc) %>%
  pivot_wider(names_from = data, values_from = per_inc)   # drop rows where either is 0

# Check the correlation
cor_value <- cor(correlation_df$AIS, correlation_df$Survey, method = "pearson")
cor_value

cor_test_result <- cor.test(correlation_df$AIS, correlation_df$Survey, method = "pearson")
cor_test_result
```

### 5.3.2 Overlapping sites:
percent site overalap between the two datasets for the downstream networks 
```{r}
overlap_df <- northland_priority_list.df %>%
  dplyr::select(Common_name, data, per_inc) %>%
  pivot_wider(names_from = data, values_from = per_inc, values_fill = 0)

# Count total sites and overlapping sites
total_sites <- nrow(overlap_df)
overlapping_sites <- overlap_df %>%
  filter(AIS > 0 & Survey > 0) %>%
  nrow()

# Percentage of overlap
percent_overlap <- (overlapping_sites / total_sites) * 100
percent_overlap
```


# 6. Relative Caulerpa pressure Vs Weighted out degree 
this code plots the relationship between relative caulerpa pressure and weighted out degree and calculates the correlation. 
Please note that weighted out degree comes from Survey data paper for entire network calculations and as such is different to weighted outdegrees calculated above. Please see Hilliam et al., 2024 https://doi.org/10.1111/1365-2664.14660 and refer to methods section in Faubel et al., 2026 (this paper) for detailed explanation. 

## 6.1 Load bi plot csv 
- load simplifed noe list with only metrcis of 
 "per_inc"   and     "wei_outdegree" 
 
```{r}
northland_biplot.df <- read_csv("data/Northland_weiOutDegree_biplot_data.csv")
```

## 6.2 plot biplot 
```{r}
northland_biplot.df2 <- northland_biplot.df %>%
  dplyr::select(nID, name, per_inc, wei_outdegree, top_10, data) %>%
  arrange(desc(wei_outdegree)) 

# dot plot with per_inc on x axis and wei_outdegree on y axis, with facet for two plots based on data
northland_bi.plot <- northland_biplot.df2 %>%
  ggplot(aes(x = per_inc, y = wei_outdegree)) +
  geom_point(aes(color = data), size = 5) +
  labs(x = "Relative Risk (%)",
       y = "Weighted Out-Degree") +
  geom_point(data = subset(wei_deg.df, top_10 == TRUE), 
              aes(x = per_inc, y = wei_outdegree),
              size = 5,
              shape = 21,                        # Circle with border
              stroke = 1.5,                      # Thickness of border
              color = "black") +                 # Border color
  scale_x_continuous(limits = c(0, 35),expand = c(0, 0)) + 
  scale_y_continuous(limits = c(0, 250), expand = c(0, 0)) + 
  scale_color_manual(name = "Node Data",values = c("#238b45", "#66c2a5", "#d98d00")) +
  theme_classic() +
  theme(axis.title = element_text(size = 14),    # Axis titles
        axis.text = element_text(size = 14),     # Axis tick labels
        legend.title = element_text(size = 14),  # Legend title
        legend.text = element_text(size = 14))   # Legend text



northland_bi.plot 
```

```{r}
# dot plot with per_inc on x axis and wei_outdegree on y axis, with facet for two plots based on data
northland_bi_test.plot <- northland_biplot.df2 %>%
  ggplot(aes(x = per_inc, y = wei_outdegree)) +
  geom_point(aes(color = data), size = 5) +
  labs(x = "Relative Risk (%)",
       y = "Weighted Out-Degree") +
  geom_point(data = subset(northland_biplot.df2, top_10 == TRUE), 
              aes(x = per_inc, y = wei_outdegree),
              size = 5,
              shape = 21,                        # Circle with border
              stroke = 1.5,                      # Thickness of border
              color = "black") +                 # Border color
  scale_x_continuous(limits = c(0, 35),expand = c(0, 0)) + 
  scale_y_continuous(limits = c(0, 250), expand = c(0, 0)) + 
  scale_color_manual(name = "Node Data",values = c("#238b45", "#66c2a5", "#d98d00")) +
  theme_classic() +
  theme(axis.title = element_text(size = 14),    # Axis titles
        axis.text = element_text(size = 14),     # Axis tick labels
        legend.title = element_text(size = 14),  # Legend title
        legend.text = element_text(size = 14))   # Legend text

northland_bi_test.plot
```


## 6.3 Correlation between Relative Caulerpa pressure and secondary spread (Weighted out degree) 

### 6.3 AIS 
```{r}
biplot_AIS.df <- northland_biplot.df %>% 
  filter(data %in% c("AIS", "Unvisited"))

# Correlation test
cor_test_result_biplot_AIS <- cor.test(biplot_AIS.df$per_inc, biplot_AIS.df$wei_outdegree, method = "pearson")
cor_test_result_biplot_AIS

```

## 6.4 survey 
```{r}
biplot_Survey.df <- northland_biplot.df %>% 
  filter(data %in% c("Survey", "Unvisited"))

# correlation test
cor_test_result_biplot_Survey <- cor.test(biplot_Survey.df$per_inc, biplot_Survey.df$wei_outdegree, method = "pearson")
cor_test_result_biplot_Survey

```


