---
  title: "S01_analysis"
output: html_document
date: "2026-03-13"
editor_options:
  chunk_output_type: console
---
  
# 1. Load packages 
```{r}
library(sf)
library(tidyverse)
library(dplyr)
library(igraph)
library(readr)
```


# 2. Load data 
```{r}
Downstream_AIS_edge.df <- read_csv("data/Downstream_AIS_edge.csv")

Downstream_AIS_nodes.df <- read_csv("data/Downstream_AIS_nodes.csv")

Downstream_survey_edge.df <- read_csv("data/Downstream_survey_edge.csv")

Downstream_survey_nodes.df <- read_csv("data/Downstream_survey_nodes.csv")

Reachability_AIS_edge.df <- read_csv("data/Reachability_AIS_edge.csv")

Reachability_AIS_nodes.df <- read_csv("data/Reachability_AIS_nodes.csv")

Reachability_survey_edge.df <- read_csv("data/Reachability_survey_edge.csv")

Reachability_survey_nodes.df <- read_csv("data/Reachability_survey_nodes.csv")
```


# 3. create igraph object and calcualte metrics 

## 3.1 Create I graph object 
adding in both the edge list and the node dataset automatically applies all of the node attributes to the igraph object 

### 3.1.1 Downstream_AIS
```{r}
head(Downstream_AIS_edge.df)

AIS_down.g <- graph_from_data_frame(Downstream_AIS_edge.df, Downstream_AIS_nodes.df , directed = TRUE)

E(AIS_down.g)$weight    <- E(AIS_down.g)$vessels

```

#### network metrics 
```{r}
# --- If you want to use weight-based distances ---
## note that this is not used within this analysis but for calculating metrics such as closeness centrality or betweenness centrality you will need
E(AIS_down.g)$strength <- E(AIS_down.g)$weight
E(AIS_down.g)$distance <- mean(E(AIS_down.g)$weight) / E(AIS_down.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(AIS_down.g)$degree    <- degree(AIS_down.g, v = V(AIS_down.g), mode = "total")
V(AIS_down.g)$indegree  <- degree(AIS_down.g, v = V(AIS_down.g), mode = "in")
V(AIS_down.g)$outdegree <- degree(AIS_down.g, v = V(AIS_down.g), mode = "out")

# --- Strength
V(AIS_down.g)$strength     <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "all", loops = TRUE)
V(AIS_down.g)$in_strength  <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "in",  loops = TRUE)
V(AIS_down.g)$out_strength <- strength(AIS_down.g, vids = V(AIS_down.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(AIS_down.g)$wei_degree     <- V(AIS_down.g)$degree * ((V(AIS_down.g)$strength / V(AIS_down.g)$degree)^0.5)
V(AIS_down.g)$wei_indegree   <- V(AIS_down.g)$indegree * ((V(AIS_down.g)$in_strength / V(AIS_down.g)$indegree)^0.5)
V(AIS_down.g)$wei_outdegree  <- V(AIS_down.g)$outdegree * ((V(AIS_down.g)$out_strength / V(AIS_down.g)$outdegree)^0.5)

```

### 3.1.2 Downstream_Survey
```{r}
head(Downstream_survey_edge.df)
head(Downstream_survey_nodes.df)

Survey_down.g <- graph_from_data_frame(Downstream_survey_edge.df, Downstream_survey_nodes.df , directed = TRUE)

E(Survey_down.g)$weight    <- E(Survey_down.g)$vessels

summary(Survey_down.g)
```

#### network metrics 
```{r}
# --- If you want to use weight-based distances ---
## note that this is not used within this analysis but for calculating metrics such as closeness centrality or betweenness centrality you will need
E(Survey_down.g)$strength <- E(Survey_down.g)$weight
E(Survey_down.g)$distance <- mean(E(Survey_down.g)$weight) / E(Survey_down.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(Survey_down.g)$degree    <- degree(Survey_down.g, v = V(Survey_down.g), mode = "total")
V(Survey_down.g)$indegree  <- degree(Survey_down.g, v = V(Survey_down.g), mode = "in")
V(Survey_down.g)$outdegree <- degree(Survey_down.g, v = V(Survey_down.g), mode = "out")

# --- Strength 
V(Survey_down.g)$strength     <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "all", loops = TRUE)
V(Survey_down.g)$in_strength  <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "in",  loops = TRUE)
V(Survey_down.g)$out_strength <- strength(Survey_down.g, vids = V(Survey_down.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(Survey_down.g)$wei_degree     <- V(Survey_down.g)$degree * ((V(Survey_down.g)$strength / V(Survey_down.g)$degree)^0.5)
V(Survey_down.g)$wei_indegree   <- V(Survey_down.g)$indegree * ((V(Survey_down.g)$in_strength / V(Survey_down.g)$indegree)^0.5)
V(Survey_down.g)$wei_outdegree  <- V(Survey_down.g)$outdegree * ((V(Survey_down.g)$out_strength / V(Survey_down.g)$outdegree)^0.5)
```

### 3.1.3 Reachability_AIS
```{r}
head(Reachability_AIS_edge.df)

AIS_reach.g <- graph_from_data_frame(Reachability_AIS_edge.df, Reachability_AIS_nodes.df , directed = TRUE)

E(AIS_reach.g)$weight  <- E(AIS_reach.g)$vessels

summary(AIS_reach.g)
```

#### network metrics 
```{r}
# --- If you want to use weight-based distances ---
## note that this is not used within this analysis but for calculating metrics such as closeness centrality or betweenness centrality you will need
E(AIS_reach.g)$strength <- E(AIS_reach.g)$weight
E(AIS_reach.g)$distance <- mean(E(AIS_reach.g)$weight) / E(AIS_reach.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(AIS_reach.g)$degree    <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "total")
V(AIS_reach.g)$indegree  <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "in")
V(AIS_reach.g)$outdegree <- degree(AIS_reach.g, v = V(AIS_reach.g), mode = "out")

# --- Strength 
V(AIS_reach.g)$strength     <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "all", loops = TRUE)
V(AIS_reach.g)$in_strength  <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "in",  loops = TRUE)
V(AIS_reach.g)$out_strength <- strength(AIS_reach.g, vids = V(AIS_reach.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(AIS_reach.g)$wei_degree     <- V(AIS_reach.g)$degree * ((V(AIS_reach.g)$strength / V(AIS_reach.g)$degree)^0.5)
V(AIS_reach.g)$wei_indegree   <- V(AIS_reach.g)$indegree * ((V(AIS_reach.g)$in_strength / V(AIS_reach.g)$indegree)^0.5)
V(AIS_reach.g)$wei_outdegree  <- V(AIS_reach.g)$outdegree * ((V(AIS_reach.g)$out_strength / V(AIS_reach.g)$outdegree)^0.5)

```

### 3.1.4 Reachability_Survey
```{r}
head(Reachability_survey_edge.df)

Survey_reach.g <- graph_from_data_frame(Reachability_survey_edge.df, Reachability_survey_nodes.df , directed = TRUE)

E(Survey_reach.g)$weight  <- E(Survey_reach.g)$vessels

summary(Survey_reach.g)
```

#### network metrics 
```{r}
# --- If you want to use weight-based distances ---
## note that this is not used within this analysis but for calculating metrics such as closeness centrality or betweenness centrality you will need
E(Survey_reach.g)$strength <- E(Survey_reach.g)$weight
E(Survey_reach.g)$distance <- mean(E(Survey_reach.g)$weight) / E(Survey_reach.g)$weight  # define your distance as in your function

# --- Degree centrality ---
V(Survey_reach.g)$degree    <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "total")
V(Survey_reach.g)$indegree  <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "in")
V(Survey_reach.g)$outdegree <- degree(Survey_reach.g, v = V(Survey_reach.g), mode = "out")

# --- Strength 
V(Survey_reach.g)$strength     <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "all", loops = TRUE)
V(Survey_reach.g)$in_strength  <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "in",  loops = TRUE)
V(Survey_reach.g)$out_strength <- strength(Survey_reach.g, vids = V(Survey_reach.g), mode = "out", loops = TRUE)

# --- Weighted degree measures ---
V(Survey_reach.g)$wei_degree     <- V(Survey_reach.g)$degree * ((V(Survey_reach.g)$strength / V(Survey_reach.g)$degree)^0.5)
V(Survey_reach.g)$wei_indegree   <- V(Survey_reach.g)$indegree * ((V(Survey_reach.g)$in_strength / V(Survey_reach.g)$indegree)^0.5)
V(Survey_reach.g)$wei_outdegree  <- V(Survey_reach.g)$outdegree * ((V(Survey_reach.g)$out_strength / V(Survey_reach.g)$outdegree)^0.5)

```

# 4 Exporting to shp 
use the igraph to shp function to export igraph object to shp file to be used for visualitation in Arcpro 

## 4.1 Down_AIS 
```{r}
igraph2shp(GBI_g_metrics,shpOut = file.path(NRC_Network_Layer_Dir, "AIS_net_GBI_16_06_2023_NZTM"),CRSVal = "EPSG:2193")
```


## 4.2 Down Survey 
```{r}
igraph2shp(GBI_g_metrics,shpOut = file.path(NRC_Network_Layer_Dir, "AIS_net_GBI_16_06_2023_NZTM"),CRSVal = "EPSG:2193")
```

## 4.3 reach AIS 
```{r}
igraph2shp(GBI_g_metrics,shpOut = file.path(NRC_Network_Layer_Dir, "AIS_net_GBI_16_06_2023_NZTM"),CRSVal = "EPSG:2193")
```

## 4.4 Reach Survey 
```{r}
igraph2shp(GBI_g_metrics,shpOut = file.path(NRC_Network_Layer_Dir, "AIS_net_GBI_16_06_2023_NZTM"),CRSVal = "EPSG:2193")
```


# 5. Priorty sites bar plot 

## 5.1 load in priorty sites csv 
```{r}
Northland_priority_list.df <- read_csv("data/Northland_dodged_barplot_data.csv")
```

## 5.2 plot dodged bar plot 
```{r}

```


## 5.3 statistics

### 5.3.1 Correlation betwen incoming vessels 
Calculating the correation between the two datasets fro incoming vessels accross the 
downstream networks
```{r}
head(Northland_priority_list.df)

# Pivot the data so that AIS and Survey per_inc values are side-by-side for each Common_name
correlation_df <- Northland_priority_list.df %>%
  dplyr::select(Common_name, data, per_inc) %>%
  pivot_wider(names_from = data, values_from = per_inc)   # drop rows where either is 0

# Check the correlation
cor_value <- cor(correlation_df$AIS, correlation_df$Survey, method = "pearson")
cor_value

cor_test_result <- cor.test(correlation_df$AIS, correlation_df$Survey, method = "pearson")
cor_test_result
```

### 5.3.2 Overlapping sites:
percent site overalap between the two datasets for the downstream networks 
```{r}
overlap_df <- Northland_priority_list.df %>%
  dplyr::select(Common_name, data, per_inc) %>%
  pivot_wider(names_from = data, values_from = per_inc, values_fill = 0)

# Count total sites and overlapping sites
total_sites <- nrow(overlap_df)
overlapping_sites <- overlap_df %>%
  filter(AIS > 0 & Survey > 0) %>%
  nrow()

# Percentage of overlap
percent_overlap <- (overlapping_sites / total_sites) * 100
percent_overlap
```


# 6. Relative Caulerpa pressure Vs Weighted out degree 
this code plots the relationship between relative caulera pressure and weighted out degree and calculates the correlation. 

## 6.1 Load bi plot csv 
- load simplifed noe list with only metrcis of 
 "per_inc"   and     "wei_outdegree" 
```{r}
Northland_biplot.df <- read_csv("data/Northland_weiOutDegree_biplot_data.csv")
```

## 6.2 plot biplot 
```{r}

```

## 6.3 Correlation between Relative Caulerpa pressure and secondary spread (Weighted out degree) 

### 6.3 AIS 
```{r}
biplot_AIS.df <- Northland_biplot.df %>% 
  filter(data %in% c("AIS", "Unvisited"))

# Correlation test
cor_test_result_biplot_AIS <- cor.test(biplot_AIS.df$per_inc, biplot_AIS.df$wei_outdegree, method = "pearson")
cor_test_result_biplot_AIS

```

## 6.4 survey 
```{r}
biplot_Survey.df <- Northland_biplot.df %>% 
  filter(data %in% c("Survey", "Unvisited"))

# correlation test
cor_test_result_biplot_Survey <- cor.test(biplot_Survey.df$per_inc, biplot_Survey.df$wei_outdegree, method = "pearson")
cor_test_result_biplot_Survey

```




